---
title: "PS11: Shapefiles"
subtitle: "STAT 133"
author: "Your Name"
format: 
  html:
    toc: true
    number-sections: true
    theme: simplex
embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(tidyverse)      # data wrangling and graphics
library(lubridate)      # for working with dates
library(sf)             # for working with geospatial vector-data
library(leaflet)        # web interactive maps
library(plotly)         # web interactive graphics
```


::: {.callout-note icon=false}

## General Instructions

-   Download the template file `ps11.qmd` to answer the questions.

-   Rename `ps11.qmd` to `ps11-first-last.qmd`, where `first` and `last` are your first and last names (e.g. `ps11-gaston-sanchez.qmd`).

-   Download the shapefile `California_Fire_Perimeters_2018_2022` from bcourses and put it in the same directory as your qmd file. It is present in the description of this assignment and under files/problem-sets/ps11/. By default, bcourses downloads the shapefile in a zipped directory. You will need to extract it before working on this problem set.

-   Render your qmd file to an HTML file and submit both your qmd and HTML files to the corresponding assignment submission in bCourses.

:::



# CalFIRE Perimeters Data 2018-2022

The geospatial data of this assignment is the __shapefile__ 
`California_Fire_Perimeters_2018_2022` located in bCourses:

<https://bcourses.berkeley.edu/courses/1551809/files/folder/problem-sets/ps11>

This is a subset of the `California_Fire_Perimeters_(+1950)` reviewed in
lecture 32 about Shapefiles (April-13). Please check the associated slides to 
know more about this data set.


## Importing shapefile data in R

As discussed in lecture, shapefiles can be imported in R with the function 
`st_read()` from the package `"sf"`.

Assuming that you have the shapefile `California_Fire_Perimeters_2018_2022` 
in your working directory, here's how to import it in R:

```{r message=FALSE}
# import shapefile
cal_perims <- st_read(
  dsn = "California_Fire_Perimeters_2018_2022",
  quiet = TRUE)
```


## A bit of Exploratory Data Analysis

The first part of this assignment involves applying basic exploratory commands 
on the fire perimeters data. While one could proceed with dplyr commands as usual, 
it is better if we create a "copy" of `cal_perims` that doesn't include the `geometry` 
column. The reason to do this is to make operations faster, otherwise computations
become slow if we are _dragging_ the `geometry` column.

Use the code cell below to create a copy of `cal_perims` called `dat_perims`, and then drop the `geometry` column from `dat_perims`. Please review the lecture on shapefiles if you need a refresher on how to drop the `geometry` column.

```{r}
# Your code here
dat_perims <- st_drop_geometry(cal_perims)
head(dat_perims)
```


Using `dat_perims`, write R commands (e.g. pipelines) for the following:

a) Find the number of fires in each year

```{r}
# Your code here
dat_perims |>
  group_by(YEAR_) |>
  summarise(num_fires = n())
```

\

b) Get the `YEAR_`, `FIRE_NAME`, `ALARM_DATE`, `CONT_DATE`, and `GIS_ACRES` of the fire with the largest area burned before containment.

```{r}
# Your code here
dat_perims |> 
  arrange(desc(GIS_ACRES)) |>
  slice(1) |>
  select(`YEAR_`, `FIRE_NAME`, `ALARM_DATE`, `CONT_DATE`, `GIS_ACRES`)

```

c) Make a bar chart to visualize the number of fires started in each month 
regardless of year (i.e. total # of fires started in Jan, total # of fires 
started in Feb, so on and so forth). TIP: you'll have to mutate a column for 
month, based on the alarm date. Make sure your graph shows names (abbreviations)
of months, and include meaningful labels (title, axis labels). Also, provide a 
concise explanation of what you see in the graph.

```{r}
# Your code here
monthly <- dat_perims |>
  mutate(month = month(ALARM_DATE)) |>
  group_by(month) |>
  summarise(num_fires = n())

ggplot(data = monthly, aes(x = month, y = num_fires)) +
  geom_col() +
  labs(x = 'Month', y = 'Number of Fires', title = 'Number of Fires per Month') +
  scale_x_continuous(breaks=seq(1:12), labels = c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec'))

```


::: {.callout-note}
# Type Your Answer in this Callout Block

It seems the summer months such as June, July, and August have the greatest number of fires. This would make sense since they are often the hottest and dryest months. 

:::

\

d) Make a plot to visualize the distribution of burned area (`GIS_ACRES`) and
provide a concise explanation of what you see in the graph.

```{r}
# Your code here
ggplot(data = dat_perims, aes(x = GIS_ACRES)) +
  geom_histogram(bins = 30)

```

::: {.callout-note}
# Type Your Answer in this Callout Block

The vast majority of fires are quite small, under 50,000 acres. However, there are a few outliers that are extremely large, up to 1 million acres.

:::

e) Obtain a table with the year, fire name, alarm date, containment date, 
and area burned of the largest fire in each year. Call this table 
`largest_fires_by_year` and display its contents.

```{r}
largest_fires_by_year <- dat_perims |>
  group_by(YEAR_) |>
  arrange(desc(GIS_ACRES)) |>
  slice(1) |>
  select(YEAR_, `FIRE_NAME`, `ALARM_DATE`, `CONT_DATE`, `GIS_ACRES`)

largest_fires_by_year
```

f) With the `largest_fires_by_year` table obtained in the preceding part, 
mutate a column `duration` with the duration between the alarm date and the
containment date. Arrange the rows of the output table by `duration` in 
descending order.

```{r}
# Your code here
largest_fires_by_year |>
  mutate(duration = CONT_DATE - ALARM_DATE) |>
  arrange(desc(duration))
```


-----


# Interactive Map

As you may recall from lecture, the polygons in `"California_Fire_Perimeters_2018_2022"`
are not expressed in long-lat format. This is an issue if you want to make 
interactive maps with leaflet. Why? Because `"leaflet"` expects all point, 
line, and polygon vector-data to be specified in longitude and latitude 
degrees using the _geodetic coordinate system_ WGS 84 (a.k.a. EPSG:4326). 

To use `leaflet()`, we need to `st_transform()` the shapefile by specifying 
the coordinate system EPSG:4326 (used by leaflet):

```{r}
# make cal_perims compatible with leaflet
cal_perims <- cal_perims |>
  st_transform(crs = 4326)
```


## Map of Large Wildfires

The goal is to visualize the fire perimeters of large fires. By "large" we
mean fires with an area burned greater than or equal to 10,000 acres.

Make a leaflet map to visualize __only large fires__ (2018-2022). Add labels to the
polygons so that when you hover over a given fire perimeter, a label is 
displayed with the name and alarm-date of the fire.

```{r}
# Your code here
large_fires <- cal_perims |>
  filter(GIS_ACRES >= 10000)

large_fires |>
  leaflet() |>
  addProviderTiles(provider = "Esri.WorldTopoMap") |> 
  addPolygons(
    color = "red",
    fillOpacity = 0.5,
    weight = 1, 
    label = ~FIRE_NAME)
```


## Map of Large Wildfires Colored By Season

Make another leaflet map to visualize __only large fires__ (2018-2022) with
the same specifications of version 1. This time, however, color code the 
fires based on the following month categories:

- `"Jan-Mar"`
- `"Apr-Jun"`
- `"Jul-Sep"`
- `"Oct-Dec"`

Your map should include a label for the colors. You will need to `mutate` a column that indicates the three-month period by using a `lubridate` function on the `ALARM_DATE` column. 

```{r}
# Your code here
seasons <- large_fires |>
  mutate(season = quarter(ALARM_DATE))

pal <- colorFactor(palette = "Set1", domain = seasons$season)

seasons |>
  leaflet() |>
  addProviderTiles(provider = "Esri.WorldTopoMap") |> 
  addPolygons(
    color = ~pal(season),
    fillOpacity = 0.5,
    weight = 1, 
    label = ~FIRE_NAME)
```


