---
title: 'project 4'
author: 'erin yang'
format:
    dashboard: 
        orientation: columns
embed-resources: True
---

```{r}
library(dplyr)
library(readr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(plotly)
```

```{r}
#| include: false
alpha <- read_csv('data/bicyclist_alpha_2014_2023.csv')

beta <- read_csv('data/bicyclist_beta_2014_2023.csv')
alpha
```


```{r}
#| include: false
alpha <- alpha %>%
  mutate(
    YEAR = lubridate::year(alpha$COLLISION_DATE),
    MONTH = lubridate::month(alpha$COLLISION_DATE),
    DAY_OF_WEEK = lubridate::wday(alpha$COLLISION_DATE, label = TRUE),
    HOUR = case_when(
        COLLISION_TIME >= 2100 ~ 21,
        COLLISION_TIME >= 2000 ~ 20,
        COLLISION_TIME >= 1900 ~ 19,
        COLLISION_TIME >= 1800 ~ 18,
        COLLISION_TIME >= 1700 ~ 17,
        COLLISION_TIME >= 1600 ~ 16,
        COLLISION_TIME >= 1500 ~ 15,
        COLLISION_TIME >= 1400 ~ 14,
        COLLISION_TIME >= 1300 ~ 13,
        COLLISION_TIME >= 1200 ~ 12,
        COLLISION_TIME >= 1100 ~ 11,
        COLLISION_TIME >= 1000 ~ 10,
        COLLISION_TIME >= 900  ~ 9,
        COLLISION_TIME >= 800  ~ 8,
        COLLISION_TIME >= 700  ~ 7,
        COLLISION_TIME >= 600  ~ 6,
        COLLISION_TIME >= 500  ~ 5,
        COLLISION_TIME >= 400  ~ 4,
        COLLISION_TIME >= 300  ~ 3,
        COLLISION_TIME >= 200  ~ 2,
        COLLISION_TIME >= 100  ~ 1,
        COLLISION_TIME >= 0    ~ 0
    )
  )


```


```{r}
#| include: false
crash <- merge(alpha, beta, by = 'CASE_ID')
head(crash)

```

# Yearly Bike Collisions

## column

Research Question: How have total injuries and deaths from bike collisions changed over the years?

```{r}
yr_deaths <- crash |>
  group_by(YEAR) |>
  summarise(sum_killed = sum(NUMBER_KILLED, na.rm = TRUE))

```

```{r}
#| TItle: Bike Collision Deaths
p1 <- ggplot(data = yr_deaths, aes(x = YEAR, y = sum_killed)) +
        geom_line() +
        geom_point() +
        labs(title = 'Total Deaths in Bike Collisions per Year')

ggplotly(p1)
```

```{r}
yr_injury <- crash |>
  group_by(YEAR) |>
  summarise(sum_injured = sum(NUMBER_INJURED, na.rm = TRUE))

```

```{r}
 p2 <- ggplot(data = yr_injury, aes(x = YEAR, y = sum_injured)) +
  geom_line() +
  geom_point() +
  labs(title = 'Total Injuries in Bike Collisions per Year')

ggplotly(p2)
```

## column 2

```{r}
yr_comb <- merge(yr_injury, yr_deaths, by = 'YEAR')

```


```{r}
#| title: Table of Yearly Bike Collision Data
yr_comb['prop_killed'] <- yr_comb$sum_killed / (yr_comb$sum_killed + yr_comb$sum_injured)

yr_comb
```

```{r}
#| component: valuebox
#| title: Average Number of Deaths per Year
list(
  icon = "bicycle",
  color = "primary",
  value = mean(yr_comb$sum_killed, na.rm = TRUE)
)
```

```{r}
p3 <- ggplot(data = yr_comb, aes(x = YEAR, y = prop_killed)) +
  geom_line() +
  geom_point() +
  labs(title = 'Proportion of Deaths in Bike Collisions per Year')

ggplotly(p3)
```

# Monthly Bike Collisions

## column

Research Question: How do the number of bike collision deaths and injuries change through the months?

```{r}
mn_deaths <- crash |>
  group_by(MONTH) |>
  summarise(sum_killed = sum(NUMBER_KILLED, na.rm = TRUE))

```



```{r}
mn_injury <- crash |>
  group_by(MONTH) |>
  summarise(sum_injured = sum(NUMBER_INJURED, na.rm = TRUE))

```



```{r}
mn_comb <- merge(mn_injury, mn_deaths, by = 'MONTH')
```


```{r}
mn_comb['prop_killed'] <- mn_comb$sum_killed / (mn_comb$sum_killed + mn_comb$sum_injured)

mn_comb
```


```{r}
p6 <- ggplot(data = mn_comb, aes(x = MONTH, y = prop_killed)) +
  geom_line() +
  geom_point() +
  labs(title = 'Proportion of Deaths in Bike Collisions per Month')

ggplotly(p6)
```

There is a greater proportion of deaths in the winter. Perhaps because environmental conditions such as light and rain are often worse.

## column

```{r}
mn_long <- mn_comb |>
  pivot_longer(
    cols = c(sum_killed, sum_injured),
    names_to = "metric",
    values_to = 'value'
  )

```


```{r}
p7 <- ggplot(data = mn_long, aes(x = MONTH, y = value)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ metric, scales = 'free_y') +
  labs(title = 'Bike Collision Deaths and Injuries by Month')

ggplotly(p7)
```

### row {height = 60%}

```{r}
#| component: valuebox
#| title: Month with the Greatest Death Rate
list(
  icon = "bicycle",
  color = "red",
  value = 'December'
)
```

### row

It seems that there are the most injuries and deaths in the fall. Perhaps this is when people bike the most often!
