---
title: 'Analysis 2'
format: 
    html:
        code-fold: true
embed-resources: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(tidytext)
library(plotly)   
```

```{r, include=FALSE}
# load sentiment datasets
bing <- read_csv('sentiment-lexicons/bing.csv')
nrc <- read_csv('sentiment-lexicons/nrc.csv')
afinn <- read_csv('sentiment-lexicons/afinn.csv')
```

```{r, include=FALSE}
# load data
sotu <- read_csv('data/sotu.csv')
```

## Net sentiment over the years

```{r}
allsotu <- sotu |> select(text)
alltokens <- unnest_tokens(tbl = allsotu, input = text, output = word)
alltokens <- anti_join(alltokens, stop_words, by = "word")

allsent = inner_join(alltokens, sentiments, by = "word")

g6 <- allsent |> 
  count(word, sentiment, name = "count", sort = TRUE) |> 
  slice_head(n = 20) |> 
  ggplot(aes(x = count, y = reorder(word, count), fill = sentiment)) +
  geom_col()

ggplotly(g6)
```

## First SOTU address contribution analysis

```{r}
fsotu <- sotu |> slice(1) |> select(text)
ftokens <- unnest_tokens(tbl = fsotu, input = text, output = word)
ftokens <- anti_join(ftokens, stop_words, by = "word")
fsent <- inner_join(ftokens, sentiments, by = "word")

g7 <- fsent |> 
  count(word, sentiment, name = "count", sort = TRUE) |> 
  slice_head(n = 20) |> 
  ggplot(aes(x = count, y = reorder(word, count), fill = sentiment)) +
  geom_col()

ggplotly(g7)
```

## Latest SOTU address contribution analysis

```{r}
lsotu <- sotu |> slice(n()) |> select(text)
ltokens <- unnest_tokens(tbl = lsotu, input = text, output = word)
ltokens <- anti_join(ltokens, stop_words, by = "word")

lsent <- inner_join(ltokens, sentiments, by = "word")

g8 <- lsent |> 
  count(word, sentiment, name = "count", sort = TRUE) |> 
  slice_head(n = 20) |> 
  ggplot(aes(x = count, y = reorder(word, count), fill = sentiment)) +
  geom_col()

ggplotly(g8)
```

## Sentiment dictionary comparison

```{r}
allafinn = inner_join(alltokens, afinn, by = "word")

g9 <- allafinn |> 
  count(word, value, name = "count", sort = TRUE) |> 
  slice_head(n = 20) |> 
  ggplot(aes(x = count, y = reorder(word, count), fill = value)) +
  labs(y = 'word', title = 'Afinn Lexicon') +
  geom_col()

ggplotly(g9)
```


```{r}
allnrc = inner_join(alltokens, nrc, by = "word")

g10 <- allnrc |> 
  count(word, sentiment, name = "count", sort = TRUE) |> 
  slice_head(n = 20) |> 
  ggplot(aes(x = count, y = reorder(word, count), fill = sentiment)) +
  labs(y = 'word', title = 'NRC Lexicon') +
  geom_col()

ggplotly(g10)
```

It's definitely easier to interpret the afinn lexicon since things are on a numeric scale between good and bad. From the NRC graph though, we can see that there are a lot of "trusting" words. I wonder why congress is 'disgust' lol