---
title: "PS1: Vectors"
subtitle: "STAT 133"
author: "Erin Yang"
format: 
  html:
    toc: true
    number-sections: true
    theme: simplex
embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

::: {.callout-note icon="false"}
## General Instructions

-   Use the template file `ps1.qmd` to answer the questions.

-   Rename this file as `ps1-first-last.qmd`, where `first` and `last` are your first and last names (e.g. `ps1-gaston-sanchez.qmd`).

-   Submit both your `qmd` and `HTML` files to the corresponding assignment submission in bCourses.
:::

# Numeric Sequences

Use the `seq()` function to create the following vectors:

a)  `1 2 3 4 5 6 7 8 . . . 18 19 20`

```{r}
# your code
a <- seq(20)
a
```

\

b)  `1.0 1.2 1.4 1.6 1.8 2.0`

```{r}
# your code
seq(1, 2, 0.2)
```

\

c)  `10 8 6 4 2 0 -2 -4 -6 -8 -10`

```{r}
# your code
seq(10, -10, -2)
```

\

d)  A sequence with the first 20 multiples of 3: `3 6 9 12 ...`. *Hint*: the `lengh.out` argument in `seq()` is your friend.

```{r}
# your code
seq(from = 3, to = (3 * 20), by = 3)
```

------------------------------------------------------------------------

# Replicates

The function `rep()` allows you to create vectors of replicated elements in a handful of different ways. The main input is a vector `x`, and you can specify arguments such as `each`, `times` and `length.out`

``` r
rep(x, each, times, length.out)
```

a)  Figure out how to use the `rep()` function and the input vector `1:3` to create the output `1 1 2 2 3 3`

```{r}
# your code
rep(1:3, each = 2)
```

\

b)  Figure out how to use the `rep()` function and the input vector `1:3` to create the output `1 2 3 1 2 3`

```{r}
# your code
rep(1:3, times = 2)
```

\

c)  Figure out how to use the `rep()` function and the input vector `1:3` to create the output `1 1 2 2 3 3 1 1 2 2 3 3`

```{r}
# your code
rep(1:3, each = 2, times = 2)
```

\

d)  Figure out how to use the `rep()` function and the input vector `1:4` to create the output `1 2 2 3 3 3 4 4 4 4`

```{r}
# your code
rep(1:4, times = 1:4)
```

------------------------------------------------------------------------

# GSW Data (season 2025)

The (raw) data for this problem has to do with some players from Golden State Warriors (2025), displayed in the table below (source: Basketball-Reference).

| number | player             | position | height | weight | experience |
|--------|--------------------|----------|--------|--------|------------|
| 30     | Stepehn Curry      | PG       | 74     | 185    | 16         |
| 10     | Jimmy Buttler      | SF       | 78     | 230    | 14         |
| 23     | Draymond Green     | PF       | 78     | 230    | 13         |
| 2      | Brandin Podziemski | SG       | 76     | 205    | 2          |
| 1      | Jonathan Kuminga   | PF       | 79     | 225    | 4          |
| 4      | Moses Moody        | SG       | 77     | 211    | 4          |
| 20     | Al Horford         | C        | 80     | 240    | 18         |
| 3      | Will Richard       | SG       | 75     | 206    | 0          |

## Creating Vectors (by hand)

Use the "combine" function `c()` to create vectors for the columns in the data table displayed above, according to the following data types.

-   `number`: integer vector

-   `player`: character vector

-   `position`: character vector

-   `height`: double (i.e. real) vector

-   `weight`: double (i.e. real) vector

-   `experience`: integer vector

```{r}
# your code
number <- c(30, 10, 23, 2, 1, 4, 20, 3)
player <- c('Stephen Curry', 'Jimmy Buttler', 'Draymond Green', 'Brandin Podziemsky', 'Jonathan Kuminga', 'Moses Moody', 'Al Horford', 'Will Richard')
position <- c('PG', 'SF', 'PF', 'SG', 'PF', 'SG', 'C', 'SG')
height <- c(74, 78, 78, 76, 79, 77, 80, 75)
weight <- c(185, 230, 230, 205, 225, 211, 240, 206)
experience <- c(16, 14, 13, 2, 4, 4, 18, 0)
```

## Subsetting (i.e. indexing) Vectors

Refer to the vectors created in the preceding subsection. Write R commands (just one single-line commands)---displaying the output---that answer the following questions:

a)  What is the weight of the heaviest player?

```{r}
# your code
max(weight)
```

<br>

b)  What is the name of the player that has a height of 80 inches\`?

```{r}
# your code
player[height == 80]
```

<br>

c)  What is the position of the player with more years of experience? *Hint*: the `which.max()` function is your friend.

```{r}
# your code
position[which.max(experience)]
```

<br>

d)  What is the number of the lightest player? *Hint*: the `which.min()` function is your friend.

```{r}
# your code
number[which.min(weight)]
```

<br>

e)  What is the average weight for those players with more than 5 years of experience?

```{r}
# your code
mean(weight[player > 5])
```

<br>

f)  How many players have a weight larger than the average (i.e. mean) weight? *Hint*: the `sum()` function is your friend.

```{r}
# your code
length(player[weight > 216.5])
```

<br>

g)  How many players have between 9 and 12 years of experience (inclusive)?

```{r}
# your code
length(experience[(experience >=9) & (experience <= 12)])
```

<br>

h)  What is the median years of experience of *Shooting Guard* (`SG`) players?

```{r}
# your code
median(experience[position == 'SG'])
```

<br>

i)  What is the median weight of those players with a position different of *Center* (`C`)?

```{r}
# your code
median(weight[position != 'C'])
```

<br>

j)  What is the first quartile (i.e. bottom 25th percentile) of years of experience among *Power Forwards* (`PF`) and *Shooting Guards* (`SG`). *Hint*: the `quantile()` function is your friend.

```{r}
# your code
quantile(experience[(position == 'PF') | (position == 'SG')], probs = 0.25)

#experience[(position == 'PF') | (position == 'SG')]
```

------------------------------------------------------------------------

# Subsetting Commands

Consider the vector `vec = c("x", "y", "a", "b")`. For each of the following parts, write at least two different subsetting commands that produce the given outputs:

a)  `"x"`

```{r}
# your code
vec <- c("x", "y", "a", "b")
vec[1]
vec[-(2:4)]
```

\

b)  `"b" "a" "y" "x"`

```{r}
# your code
vec[4:1]
rev(vec)
```

\

c)  `"b" "b" "b"`

```{r}
# your code
rep(vec[4], times = 3)
rep(vec[4], each = 3)
```

\

d)  `"x" "y"`

```{r}
# your code
vec[1:2]
rep(vec[1:2], times = 1)
```

------------------------------------------------------------------------

# Technical questions about vectors

Consider the following two vectors `x` and `y`:

```{r}
x <- c(1, 2, 3, 4, 5)

y <- c("a", "b", "c", "d", "e")

y[x <= 2L]
```

In the following parts, indicate True or False for whether the given command involves: a) coercion, and/or b) vectorization, and/or c) recycling.

1)  `y[x <= 2L]`

    a)  coercion True
    b)  vectorization True
    c)  recycling True

2)  `y[x[5]]`

    a)  coercion True
    b)  vectorization False
    c)  recycling False

3)  `x[y != 'c']`

    a)  coercion False
    b)  vectorization True
    c)  recycling True

4)  `y[x/x]`

    a)  coercion True
    b)  vectorization True
    c)  recycling False

------------------------------------------------------------------------

# File Paths

Consider the following file structure depicted below. Note that the first forward slash `/` indicates the root directory.

```         
/
  project/
    README.md
    data/
      rawdata/
        rawdata.csv
        rawdata-dictionary.md
      cleandata/
        cleandata.csv
        cleandata-dictionary.md
    code/
      functions.R
      clean-data-script.R
```

\

a)  What is the absolute path to the folder `cleandata/`?

```{bash eval = FALSE}
/project/data/cleandata/
```

\

b)  What is the relative path from `functions.R` to `cleandata.csv`?

```{bash eval = FALSE}
../data/cleandata/cleandata.csv
```

\

c)  What is the relative path from `rawdata.csv` to `clean-data-script.R`?

```{bash eval = FALSE}
../../code/clean-data-script.R
```

\

d)  Assume that your current directory is the `cleandata/` directory. Write the Unix command that lists the contents in the `project/` directory.

```{bash eval = FALSE}
ls ../../
```

\

e)  Assume that your current directory is the `data/` directory. Write the Unix command that makes a new directory called `output/` located at the same level of `data/` and `code/`.

```{bash eval = FALSE}
mkdir ../output/
```