---
title: my first slide deck
author: Erin Yang
format:
    revealjs:
        embed-resources: true
---

# title slide 1

# title slide 2

## slide w actual content {background-color='blue'}

have you ever thought about your own mortality?

### ooga booga

```{r}
print('wheres the code bro')
```

u gotta do #| echo: true
```{r}
#| echo: true

wowza <- function(a){
    return(paste(a, '!!!!!!!!!!!!'))
}

wowza('wowza')
```

# title slide 3 {background-color='#afeeee'}


## that feeling when u

`'r a code cell'`

`2 + 2`

to return the output of the code, do 'r ____'

`r 2 + 2`

## expandable code fold!

```{r}
#| echo: true
#| code-fold: true
print('surprise!')
# it wasn't much of a surprise was it...
```

::: {.incremental}
- bullet points
- more bullet points
- MORE BULLET POINTS!!!
:::

## uhh stanford thing?


```{r}
#| echo: true
#| code-fold: true
mascot <- 'bear'

if (mascot == 'bear'){
    print('Berkeley!!')
} else if (mascot == 'tree'){
    print('stanford boo')
} else{
    print('ur only options are bear or tree')
}
```

# thank u for watching my presentation!!! {background-gradient='radial-gradient(#afeeee, #f27bffff)'}