---
title: "PS2: Matrices and Lists"
subtitle: "STAT 133"
author: "Erin Yang"
format: 
  html:
    toc: true
    number-sections: true
    theme: simplex
embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

::: {.callout-note icon="false"}
## General Instructions

-   Use the template file `ps2-template.qmd` to answer the questions.

-   Rename this file as `ps2-first-last.qmd`, where `first` and `last` are your first and last names (e.g. `ps2-gaston-sanchez.qmd`).

-   Submit both your qmd and HTML files to the corresponding assignment submission in bCourses.
:::

# Subsetting Matrices

Consider the matrix `mat = matrix(letters[1:15], nrow = 3, ncol = 5)` for the following questions.

```{r}
mat = matrix(letters[1:15], nrow = 3, ncol = 5)
mat
```

For each of the following parts, write a subsetting command that produce the given outputs:

a)  `"e"`

```{r}
# your command
mat[2, 2]
```

\

b)  Matrix 2x2 with `"e", "h"` as a first row and `"f", "i"` as a second row

```{r}
# your command
mat[2:3, 2:3]
```

\

c)  Vector `"a" "g" "m"`.

```{r}
# your command
mat[1,c(1, 3, 5)]
```

\

d)  Matrix 3x3 in which the original 2nd and 4th columns are excluded.

```{r}
# your command
mat[, c(1, 3, 5)]
```

\

e)  Matrix 3x5 in which the order of the rows is reversed (i.e. row 1 becomes row 3, row 2 becomes row 2, row 3 becomes row 1).

```{r}
# your command
mat[c(3, 2, 1),]
```

------------------------------------------------------------------------

# Creating a List

Consider an R list `cass` that when printed it gets displayed as follows:

```         
$name
    first    middle      last 
"Cassian"   "Jeron"   "Andor" 

$skills
[1] "captain" "pilot"   "lier"    "shooter"

$cargo
    X  Y
a 101 90
b 201 80
c 301 70

$specs
$specs$species
[1] "human"

$specs$gender
[1] "male"

$specs$height
[1] 1.78
```

Write code to create the list `cass`, and display this object (i.e. print it). Notice that `name` is a named character vector, `skills` is a character vector with unnamed elements, `cargo` is a matrix with row and column names, and `specs` is a list with 3 named elements.

```{r}
# your code

name = c(first = 'Cassian', middle = 'Jeron', last = 'Andor')
skills = c("captain", "pilot",   "lier", "shooter")
cargo = matrix(data = c(101, 201, 301, 90, 80, 70), nrow = 3, ncol = 2)
rownames(cargo) = c('a', 'b', 'c')
colnames(cargo) = c('X', 'Y')
specs = list('species' = 'human', 'gender' = 'male', 'height' = '1.78')

cass = list('name' = name, 'skills' = skills, 'cargo' = cargo, 'specs' = specs)
cass
```

------------------------------------------------------------------------

# Subsetting Lists

Consider the following list:

```{r}
list_example = list("letter" = c("A", "B", "C"), "value" = c(10, 50, 20))
list_example['letter']
list_example[["letter"]]
```

(a) What is the difference between `list_example["letter"]` and `list_example[["letter"]]`? Explain.

**Your Answer:**

one bracket means to return the vector "letter" within list_example. two brackets means to return the vector c("A", "B", "C") that is assigned to 'letter'. this does not return the vector name "letter" itself.\

(b) What does `list_example$letter[3]` return? Why?

**Your Answer:**

This returns a list with the item "C" because the brackets return a list as well.\

(c) What does the command `list_example$value = NULL` do? Explain.

**Your Answer:**

This would not return a value, but it would change all items in the vector "value" to NULL.\

(d) What does the command `list_example$lgl = c(TRUE, FALSE, TRUE)` do? Explain.

**Your Answer:**

this adds a new item to the list, which is a list titled lgl that contains the values TRUE, FALSE, TRUE.

\

------------------------------------------------------------------------

# Superheroes

Consider the following image containing data of a handful of superheroes.

::: callout-important
Refer to the file `ps2.html` to see the image with the superheroes data.
:::

## Creating Vectors

Create vectors for the columns in the data table displayed above, according to the following data types. If there are missing values, encode them as `NA`.

-   `name`: character vector (name of superhero)

-   `age`: integer vector (years)

-   `height`: real (i.e. double) vector (meters)

-   `weight`: real (i.e. double) vector (kg)

-   `strength`: integer vector

```{r}
# your code
name <- c('Superman', 'Hulk', 'Thor', 'Batman', 'Ironman')
age <- c(18, 35, NA, 40, 28)
height <- c(1.90, 5.02, 1.98, 1.88, 1.85)
weight <- c(107, 1089, 110, 95, 102)
strength <- c(3, 3, 3, 1, 2)
```

------------------------------------------------------------------------

# Matrix of Superheros

## Creating a Matrix

Use the vectors created in the preceding part to create a matrix `HEROES`. This matrix should have dimensions: 5 rows, and 4 columns (age, height, weight, and strength). Also, give names to both the rows and columns of `HEROES` according to figure 1; and display this matrix.

```{r}
# your code
HEROES <- matrix(data = c(age, height, weight, strength), ncol = 4)

rownames(HEROES) <- name
colnames(HEROES) <- c('age', 'height', 'weight', 'strength')
HEROES
```

## Subscripting (aka Indexing, Slicing) a Matrix

Refer to the matrix `HEROES`.

a)  Using `HEROES`, write a command that gives you the first 3 rows, and columns `height` and `weight`.

```{r}
# your code
HEROES[1:3, 2:3]
```

\

b)  Using `HEROES`, write a command that gives you the following output (see below): age values arranged in decreasing order. *Hint*: `sort()` is your friend.

```         
  Batman     Hulk  Ironman Superman 
      40       35       28       18
```

```{r}
# your code
sort(HEROES[,1], decreasing = TRUE)
```

\

c)  Using `HEROES`, write a command that gives you the name of the youngest superhero. *Hint*: `which.min()` is your friend.

```{r}
# your code
which.min(HEROES[,1])
```

\

d)  Using `HEROES`, write a command that gives you the number of superheroes that have a strength level less than or equal to 2. *Hint*: the function `sum()` is your friend.

```{r}
# your code
sum(HEROES[,4] <= 2)
```

\

e)  Using `HEROES`, write a command that arranges its content by `weight` in decreasing order. *Hint*: the function `order()` is your friend.

```{r}
# your code
HEROES[order(HEROES[,'weight'], decreasing = TRUE),]

```

\

f)  Add a new row to `HEROES` with the data of Aquaman given below. *Hint*: the `rbind()` function is your friend. Display the updated `HEROES` matrix. To clarify: `HEROES` should have 6 rows (Aquaman's data in row 6), and it should also include Aquaman's name in the row names.

-   name: Aquaman
-   age: 35
-   height: 1.85
-   weight: 147
-   strength: 3

```{r}
# your code
Aquaman <- c(35, 1.85, 147, 3)
HEROES <- rbind(HEROES, Aquaman)
HEROES
```

\

g)  Add a column `ratio` to `HEROES`. The values of `ratio` should be the ratio of weight over height. *Hint*: the `cbind()` function is your friend. Display the updated `HEROES` matrix. To clarify: `HEROES` should have 6 rows and 5 columns.

```{r}
# your code
ratio <- c(HEROES[,'weight']/HEROES[,'height'])
HEROES <- cbind(HEROES, ratio)
HEROES
```

\

------------------------------------------------------------------------

# List of Superheroes

Create a list `heroes` containing the vectors created in problem 4. This list should have five elements (one vector per element). Make sure each element of the list has the corresponding name so that you can refer to them by invoking either `heroes$name` or `heroes[["name"]]`. Display your `heroes` list.

```{r}
# your code
name <- c('Superman', 'Hulk', 'Thor', 'Batman', 'Ironman')
age <- c(18, 35, NA, 40, 28)
height <- c(1.90, 5.02, 1.98, 1.88, 1.85)
weight <- c(107, 1089, 110, 95, 102)
strength <- c(3, 3, 3, 1, 2)
heroes <- list('name' = name, 'age' = age, 'height' = height, 'weight' = weight, 'strength' = strength)
heroes
```

## Subscripting (aka Indexing, Slicing) a List

**Refer to the list `heroes`**. Write R commands---displaying the output---that answer the following questions. Answer each part in its individual code chunk.

a)  What is the name of the superhero with smallest strength level? *Hint*: the function `which.min()` is your friend.

```{r}
# your code
heroes$name[which.min(heroes$strength)]
```

\

b)  How many superheroes have an unknown `age` value? *Hint*: the function `is.na()` is your friend.

```{r}
# your code
sum(is.na(heroes$age))
```

\

c)  What is the median age of superheroes (ignore missing values)?

```{r}
# your code
median(heroes$age[-3])
```

\

d)  What is the height of the superhero with the smallest weight value? *Hint*: the function `which.min()` is your friend.

```{r}
# your code
heroes$height[which.min(heroes$weight)]
```

\

e)  How many superheroes have height values between 1.9 and 2.5 meters (inclusive)? *Hint*: the function `sum()` is your friend.

```{r}
# your code
sum(heroes$height >= 1.9 & heroes$height <= 2.5)
```

\

f)  Sort the superheroes' names arranged---from smallest to largest---by their weight/height ("weight over height") values. *Hint*: the function `order()` is your friend.

```{r}
# your code
heroes$name[order(heroes$weight / heroes$height)]
```